### 2019-05-03 papers/haskell19/4-generators.tex
Work on case studies/related work

\begin{code}
genEval  :: forall f a dot  ( BArbitrary1 f
                            , HasTerminals f
                            , Algebra f a) => BGen a
genEval d = eval <$> bounded at (Fix f) d
\end{code} %$

@@

\begin{code}
genHtmlPats :: BGen Html
genHtmlPats = genEval at Html_simplify

genValidHtml :: BGen Html
genValidHtml = genEval  at ValidHtml
\end{code}

>>>>

\begin{code}
genRep  :: forall f a dot  ( BArbitrary1 f
                           , HasTerminals f
                           , Algebra f a) => BGen a
genRep d = eval <$> bounded at (Fix f) d
\end{code} %$

@@

\begin{code}
genHtmlPats :: BGen Html
genHtmlPats = genRep at Html_simplify

genValidHtml :: BGen Html
genValidHtml = genRep  at ValidHtml
\end{code}

### 2019-05-03 papers/haskell19/5-typelevel.tex
Work on case studies/related work

And simply obtain a concrete generator by refering to |Html_Spec| when using
|genEval|:

\begin{code}
genHtml :: BGen Html
genHtml = genEval at Html_Spec
\end{code}

@@

\begin{code}
genIntTree :: BGen (Tree Int)
genIntTree = genEval at (Tree_Spec apply' Int)
\end{code}

>>>>

And simply obtain a concrete generator by refering to |Html_Spec| when using
|genRep|:

\begin{code}
genHtml :: BGen Html
genHtml = genRep at Html_Spec
\end{code}

@@

\begin{code}
genIntTree :: BGen (Tree Int)
genIntTree = genRep at (Tree_Spec apply' Int)
\end{code}

### 2019-05-03 papers/haskell19/6-casestudies.tex
Work on case studies/related work

The random generation framework presented throughout this work allows us to
write extensible generators in a very concise way.
%
However, this expresiveness comes attached to a perceptible runtime overhead,
primarly inherited from the using the \`a la Carte technique.
%
In this section we evaluate the implicit cost of composing generators using
three real-world case studies.

As we have shown in Section \ref{sec:generators}, the random generation
process we propose in this paper can be seen as having two phases.
%
First, generate random values from the representation types used to specify the
shape of our data; and then we use their algebras to translate them to the
corresponding values of our target data types.
%
In particular, the tranformation step is expected to pattern match repeatedly
against the |InL| and |InR| constructors of the |oplus| operators when
traversing each construction injection.
%
Because of this, in general we expect a performance impact with respect to
manually-written concrete generators.

As recently analyzed by \citeauthor{KiriyamaOptimizingDTC}, this slowdown is
expected to be linear in the depth of our representation type
\cite{KiriyamaOptimizingDTC}.
%
Hence, one can drastically reduce the runtime overhead by associating each
|oplus| operator in a balanced fashion.
%
So, for instance, instead of writing |(f oplus'' g oplus'' h oplus'' i)|, which
is implicitly parsed as |(f oplus'' (g oplus'' (h oplus'' i)))|; we can
associate constructions as |((f oplus'' g) oplus'' (h oplus'' i))|, thus
reducing the depth of our representation from four to three levels and, in
general, from a $\mathcal{O}(n)$ to a $\mathcal{O}(log(n))$ complexity in the
runtime overhead, where $n$ is the amount of constructions under consideration.

In this work, we analyzed the performance of generating random values using
three case studies:
%
\begin{inparaenum}[(i)]
\item Red-Black Trees, inspired by Okasaki's formulation,
\item Lisp S-expressions, inspired by the package \emph{hs-zuramaru}, and
\item HTML expressions, inspired by the \emph{html} package, which follows the
  same structure as our motivating |Html| example.
\end{inparaenum}
%
The expanse of each case study can be summarized as follows:

\begin{table}[H]
  \begin{tabular}{||c||c||c||c||c||}
    \hline
    Case Study & \#|Con| & \#|Pat| & \#|Fun| & Total Constructions \\ \hline
    \hline
    RBT        & 2      & 6      & 5      & 13                  \\ \hline
    SExp       & 6      & 9      & -      & 15                  \\ \hline
    HTML       & 4      & -      & 132    & 136                 \\ \hline
  \end{tabular}
\end{table}

We benchmarked the execution time required to generate and fully evaluate 10000
random values corresponding to each case study, comparing both manually-written
concrete generators and modular generators obtained using our approach.
%
For this purpose, we used the \emph{Criterion} benchmarking tool for Haskell,
and limited the maximum depth of the generated values to five levels.
%
Additionally, our modular generators were tested using both linear and balanced
generation specifications.
%
Figure \ref{fig:times} illustrates the relative execution time of each case
study, with respect to each manually-written random generator---we encourage the
reader to obtain a colored version of this work.

\begin{figure}[b]
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [ height=6cm
      , width=9cm
      , ybar
      , ymajorgrids=true
      , ytick={0,1,...,4}
      , ymax=4
      , ymin=0.5
      , enlargelimits=0.25
      , enlarge x limits=0.25
      , /pgf/bar width=17pt
      , legend style={
        at={(0.5,-0.15)},
        anchor=north,legend columns=3}
      , legend image code/.code={%
        \draw[#1, draw=none] (0cm,-0.1cm) rectangle (0.3cm,0.2cm); }
      , ylabel={Relative Generation Time}
      , symbolic x coords={RBT, SExp, HTML}
      , xtick=data
      % , nodes near coords
      , nodes near coords align={vertical}
      ]
      \addplot[fill, red!30]   coordinates {(SExp,1)    (RBT,1)    (HTML,1)};
      \addplot[fill, green!30] coordinates {(SExp,2.40) (RBT,3.57) (HTML,4.27)};
      \addplot[fill, blue!30]  coordinates {(SExp,1.60) (RBT,2.52) (HTML,1.55)};
      \legend{Manual, Modular (lin), Modular (bal)}
    \end{axis}
  \end{tikzpicture}
  \caption{Generation time comparison between concrete and modular generators.}
  \label{fig:times}
\end{figure}

As it can be observed, our approach suffers from a noticeable runtime overhead
when using linearly defined representations, specially when considering the HTML
case study with a large amount of construction involved at the generation
process.
%
However, we found that, by balancing our representation types the generation
performance improves dramatically.
%
At the light of these improvements in the performance, we included in our tool
an additional type-level computation that automatically balances our
representations in order to reduce the generation overhead.

@@

On the other hand, it has been argued that the generation time is often not
substantial with respect to the rest of the testing process, especially when
testing complex properties over monadic code, as well as using randomly
generated values for penetration testing.

All in all, we consider that these results are fairly encouraging, given that
the flexibility obtained from using our compositional approach does not produce
severe slowdowns in whem generating random values in practice.

>>>>

The random generation framework presented throughout this work allows us to
write extensible generators in a very concise way.
%
However, this expresiveness comes attached to a perceptible runtime overhead,
primarly inherited from the using the Data Types \`a la Carte---a technique
which is not often scrutinized for performance.
%
In this section, we evaluate the implicit cost of composing generators using
three real-world case studies.

As we have shown in Section \ref{sec:generators}, the random generation
process we propose in this paper can be seen as having two phases.
%
First, generate random values from the representation types used to specify the
shape of our data; and then we use their algebras to translate them to the
corresponding values of our target data types.
%
In particular, the tranformation step is expected to pattern match repeatedly
against the |InL| and |InR| constructors of the |oplus| operators when
traversing each construction injection.
%
Because of this, in general, we expect a performance impact with respect to
manually-written concrete generators.

As recently analyzed by \citeauthor{KiriyamaOptimizingDTC}, this slowdown is
expected to be linear in the depth of our representation type
\cite{KiriyamaOptimizingDTC}.
%
In this light, one can drastically reduce the runtime overhead by associating
each |oplus| operator in a balanced fashion.
%
So, for instance, instead of writing |(f oplus'' g oplus'' h oplus'' i)|, which
is implicitly parsed as |(f oplus'' (g oplus'' (h oplus'' i)))|; we can
associate constructions as |((f oplus'' g) oplus'' (h oplus'' i))|, thus
reducing the depth of our representation from four to three levels and, in
general, from a $\mathcal{O}(n)$ to a $\mathcal{O}(log(n))$ complexity in the
runtime overhead, where $n$ is the amount of constructions under consideration.

Then, we analyzed the performance of generating random values using three case
studies:
%
\begin{inparaenum}[(i)]
\item Red-Black Trees (RBT), inspired by Okasaki's formulation
  \cite{okasaki1999red},
\item Lisp S-expressions (SExp), inspired by the package
  \emph{hs-zuramaru}\footnote{\url{http://hackage.haskell.org/package/zuramaru}},
  and
\item HTML expressions (HTML), inspired by the \emph{html}
  package\footnote{\url{http://hackage.haskell.org/package/html}}, which follows
  the same structure as our motivating |Html| example.
\end{inparaenum}
%
The expanse of each case study can be summarized as follows:

\begin{table}[H]
  \begin{tabular}{||c||c||c||c||c||}
    \hline
    Case Study & \#|Con| & \#|Fun| & \#|Pat| & Total Constructions \\ \hline
    \hline
    RBT        & 2       & 5       & 6       & 13                  \\ \hline
    SExp       & 6       & -       & 9       & 15                  \\ \hline
    HTML       & 4       & 132     & -       & 136                 \\ \hline
  \end{tabular}
\end{table}

These case studies provide a good combination of constructions, interface
functions and patterns, and cover from smaller to larger numbers of
constructions.

We benchmarked the execution time required to generate and fully evaluate 10000
random values corresponding to each case study, comparing both manually-written
concrete generators and those obtained using our modular approach.
%
For this purpose, we used the \emph{Criterion} \cite{criterion} benchmarking
tool for Haskell, and limited the maximum depth of the generated values to five
levels.
%
Additionally, our modular generators were tested using both linear and balanced
generation specifications.
%
Figure \ref{fig:times} illustrates the relative execution time of each case
study, normalized to their corresponding manually-written counterparts---we
encourage the reader to obtain a colored version of this work.

\begin{figure}[t]
  \centering
  \begin{tikzpicture}
    \begin{axis}
      [ height=6cm
      , width=9cm
      , ybar
      , ymajorgrids=true
      , ytick={0,1,...,4}
      , ymax=4
      , ymin=0.5
      , enlargelimits=0.25
      , enlarge x limits=0.25
      , /pgf/bar width=17pt
      , legend style={
        at={(0.5,-0.15)},
        anchor=north,legend columns=3}
      , legend image code/.code={%
        \draw[#1, draw=none] (0cm,-0.1cm) rectangle (0.3cm,0.2cm); }
      , ylabel={Relative Generation Time}
      , symbolic x coords={RBT, SExp, HTML}
      , xtick=data
      % , nodes near coords
      , nodes near coords align={vertical}
      ]
      \addplot[fill, red!30]   coordinates {(SExp,1)    (RBT,1)    (HTML,1)};
      \addplot[fill, green!30] coordinates {(SExp,2.40) (RBT,3.57) (HTML,4.27)};
      \addplot[fill, blue!30]  coordinates {(SExp,1.60) (RBT,2.52) (HTML,1.55)};
      \legend{Manual, Modular (lin), Modular (bal)}
    \end{axis}
  \end{tikzpicture}
  \caption{Generation time comparison between concrete and modular generators.}
  \label{fig:times}
\end{figure}

As it can be observed, our approach suffers from a noticeable runtime overhead
when using linearly defined representations, specially when considering the HTML
case study, which spans a large amount of constructions in the generation
process.
%
However, we found that, by balancing our representation types the generation
performance improves dramatically.
%
At the light of these improvements, our tool includes an additional type-level
computation that automatically balances our representations in order to reduce
the generation overhead.

@@

On the other hand, it has been argued that the generation time is often not
substantial with respect to the rest of the testing process, especially when
testing complex properties over monadic code, as well as using randomly
generated values for penetration testing
\cite{DBLP:conf/haskell/MistaRH18,grieco2017}.

All in all, we consider that these results are fairly encouraging, given that
the flexibility obtained from using our compositional approach does not produce
severe slowdowns in when generating random values in practice.

### 2019-05-03 papers/haskell19/7-related.tex
Work on case studies/related work

\section{Related Work}
\label{sec:relatedwork}

Nullam eu ante vel est convallis dignissim. Fusce suscipit, wisi nec facilisis
facilisis, est dui fermentum leo, quis tempor ligula erat quis odio. Nunc porta
vulputate tellus. Nunc rutrum turpis sed pede. Sed bibendum. Aliquam posuere.
Nunc aliquet, augue nec adipiscing interdum, lacus tellus malesuada massa, quis
varius mi purus non odio. Pellentesque condimentum, magna ut suscipit hendrerit,
ipsum augue ornare nulla, non luctus diam neque sit amet urna. Curabitur
vulputate vestibulum lorem. Fusce sagittis, libero non molestie mollis, magna
orci ultrices dolor, at vulputate neque nulla lacinia eros. Sed id ligula quis
est convallis tempor. Curabitur lacinia pulvinar nibh. Nam a sapien.

>>>>

\section{Related Work}
\label{sec:related}

\citeauthor{SwierstraDTC} \cite{SwierstraDTC} proposes the Data Types \`a la
Carte approach for building extensible data types and unified interpreters as a
solution for the expression problem coined by \citeauthor{wadler1998expression}
\cite{wadler1998expression}.
%
In this work, we take ideas from this approach and extend them to work in the
scope of random data generation, where other parameters come into play apart
from just combining construction, e.g., generation frequency and terminal
constructions.

From the practical point of view, \citeauthor{KiriyamaOptimizingDTC} propose an
optimization mechanism for Data Types \`a Carte, where a concrete data type has
to be derived for each different composition of constructions defined by the
user \cite{KiriyamaOptimizingDTC}.
%
This solution avoids much of the runtime overhead introduced when internally
pattern matching against sequencies of |InL| and |InR| data constructors.
%
However, we see this solution as not entirely compositional, since we still need
to rely on Template Haskell to derive the machinery for each specialized data
type.
%
In our particular setting, we found that our solution has a fairly acceptable
overhead by automatically balancing our representation types.

In the past few years, there has been a bloom of derivation tools for
synthesizing random generators.
%
\emph{Feat} provides a mechanism to uniformly generate values from a given data
type \cite{DuregardJW12}.
%
It works by enumerating all the possible values of such type, so that sampling
uniformly from it simply becomes sampling uniformly from the a finite preffix of
natural numbers.

\emph{MegaDeTH} is a simple derivation tool that derives generators solely based
on their types, and paying no attention whatsoever on the generation frequency
of each data constructor.
%
As a result, its automatically derived generators has been shown to be skewed to
generating small values\cite{DBLP:conf/haskell/MistaRH18}.

On the other hand, \emph{DRAGEN} is a tool that synthesizes optimized
generators, tuning their generation frequencies using a simulation-based
optimization process parametrized by the distribution of values desired by the
user \cite{DBLP:conf/haskell/MistaRH18}.
%
This simulation is based on the theory of \emph{branching processes}, which
models the growth and extinction of populations across succesive generations.
%
In this setting, populations consist of randomly generated data constructors,
where generations correspond to each level of the generated values.
%
This tool has shown to improve the code coverage obtained over our codebase,
with respect to other fully-automated generators derivation mechanism.
%
In a recent work, \citeauthor{Mista2019GeneratingRS} extend this approach to
generate random values considering also the other sources of structural
information covered here, namely abstract interfaces and function pattern
machings.
%
The authors of this work focus on the generation model problem, extending the
theory of branching processes in order to obtain sound predictions about
distributions of random values considering these new kinds of constructions.
%
Using this extension, they show that using extra information when generating
random values is extremely valuable, in particular under situations like the
ones described in Section \ref{sec:sources}, where the usual derivation
approaches fail to synthesize useful generators due to a lack of structural
information.

This paper tackles the representation problem, demonstrating how a compositional
generation process can be effectively implemented and automated in Haskell using
advanced type-level features.
%
Since none of the aforementioned tools are designed for composability, we
consider that our ideas could perhaps be applied to improve these tools in the
future.
